The Evolution of Phenotypically Invariant Genetic Networks

### 2017-08-29 kalman_walk_paper.tex
edited system description, qgen, examples, speciation and more.

\section*{Gene Networks as Linear Dynamical Systems}
%
\plr{Below we mix general language with language specific to GRNs, like switching between "internal state" and "transcription factor concentration.  
Rather than keep making comments like "(or some other system)",
we could say up top that this applies to other situations, but to make it concrete we'll talk about regulatory networks.}
%
%  \jss{Organisms' phenotypes are constructred by gene by gene by environment interactions. Here we simply define the phenotype to be the organismal temporal molecular dynamics directly under natural selection. The \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise relevant for survival. 
%   Thus we say that some function $\phi(t)$ is a phenotype where, 
%  \begin{align}
%    \phi(t) = \int_{0}^{\infty} h(t) u(t) dt  ,
%  \end{align}
%  and $h(t)$ is the \emph{impulse response} of the system and $u(t)$ is the \emph{input} function, both functions of time $t$. The input can be interpreted as the environment, as initial conditions, or otherwise, depending on the biological specifics under study. The phenotype is a convolution of both the impulse response and the input, as allowed by our assumption of linearity. 
%
%  Essentially the phenotype $\phi(t)$ is a consequence of an organism's specific gene by gene interactions, given by $h(t)$, reacting further with the local environment, given by $u(t)$. 
%
%  We describe the impulse response as, 
%  \begin{align}
%    h(t) = C e^{A t} B
%  \end{align}
%  where $A$ is a gene regulatory network (although this model can be generalized to other biological networks) -- a square matrix, and $B$ filters and translates the input to the system. The form of $B$ determines precisely how the state of the external environment influences the internal gene network. $C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection. 
%
%  Generally $A$ can be any real $n \times n$ matrix, $B$ any $n \times \ell$, and $C$ any $\ell \times n$ dimensional matrix. Each $i$th row of $A$ describes the \emph{cis}-regulatory module for gene $i$, and each $j$th entry, the specific regulatory influence of gene $j$ on gene $i$. 
%
%  Although $\phi(t)$ describes the phenotype given an input, $h(t)$ describes the phenotype subject only to an impulse -- an input present initially and absent immediately thereafter. Typically, a system $\Sigma$, is defined as,
%  \begin{align}
%    \Sigma = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
%  \end{align}
%  Variables have the same identities as described above and $\kappa(t)$ is a vector of molecule concentrations at time $t$. Therefore the molecular concentrations at a specific time are completely determined by the input and gene by gene interactions. Lastly, a portion and/or combination of these molecules, $\phi(t)$, are ``observed'' by selection (this is in constrast to $\kappa(t)$ -- the \emph{kryptotype} -- as it is ``hidden'' from direct selection).
%}
Organisms' phenotypes are constructred by gene by gene by environment interactions. Here we simply define the \emph{phenotype} to be the temporal molecular dynamics directly under natural selection. The \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.

@@

Finally, we can write the phenotype as a convolution of the system organization and the environment,
  \begin{align}
    \phi(t) = C e^{A t} \kappa(0) + \int_{0}^{t} C e^{A (t-s)} B u(s) ds ,
  \end{align}
where we refer to $h(t) := Ce^{A t}B$ as the system's \emph{impulse response}. 
%We will now lay out the model in more general terms.
%Suppose that the \emph{internal state} of the system
%is parameterized by the concentrations of a collection of $n$ molecular species,
%$S_1, \ldots, S_n$,
%and the vector of concentrations at time $t$ we denote $\kappa(t)=(\kappa_1(t),\ldots,\kappa_n(t))$.
%There are also $m$ ``input'' species, whose concentrations are determined
%exogenously to the system,
%and are denoted $u(t) = (u_1(t),\ldots,u_m(t))$,
%and $\ell$ ``output'' species, whose concentrations are denoted
%$\phi(t) = (\phi_1(t),\ldots,\phi_\ell(t))$.
%The output is merely a linear function of the internal state:
%\begin{align*}
%    \phi_i(t) = \sum_j C_{ij} \kappa_i(t).
%\end{align*}
%Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype},
%and sometimes in contrast refer to $\kappa$ as the \emph{kryptotype},
%as it is ``hidden'' from direct selection.
%The rate at which the $i^\text{th}$ species is produced
%is a weighted sum of the concentrations of the other species
%as well as the input:
%\begin{align*}
%    \ddt \kappa_i(t) = \sum_j A_{ij} \kappa_j(t) + \sum_r B_{ir} u_r(t) .
%\end{align*}
%In matrix notation, this is written more concisely as
%\begin{align} \label{eqn:lti_system}
%    \ddt \kappa(t) &= A \kappa(t) + B u(t) \\
%    \phi(t) &= C \kappa(t) .
%\end{align}
 \begin{example}[Oscillating Gene Network: Cell Cycle Control]\label{ex:oscillator}
    Cellular division is governed by many different processes, however it is thought that its rhythm is partially controlled by oscillating gene transcription \citep{orlando2008global}. Here we consider a simplified model of oscillating gene transcription.

@@

  \section*{Phenotype-equivalent gene networks}

@@

Gene regulatory networks with identical phenotypes do not necessarily have identical kryptotypes.
Any linear and minimal system (a gene network) 
-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes 
-- has an identical impulse response and therefore an identical phenotype given an identical input $u(t)$, up to a change of coordinates. 
\plr{"has an identical phenotype" is not what you mean to say here}
\js{edited -- does that work?}
  \begin{equation}
    \begin{aligned}
      h(t) &= C e^{A t} B 
      = C V^{-1} V e^{A t} V^{-1} V B \\
      &= C V^{-1} e^{V A V^{-1} t} V B 
      = \bar{C} e^{\bar{A} t} \bar{B}
    \end{aligned}
  \end{equation}
\plr{what are we calling these -- not "biological system" -- that's too broad}\js{I see your point, but I don't want to bog down the paper by constantly qualifying statements -- it seemsnatural to refer to biological systems, since these are, afterall dynamical systems.}
Two biological systems, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, 
have the same phenotype, for all possible inputs, if they are related by a change of coordinates. 
\plr{is this the place to say "and only if, in the minimal dimension"?}

@@

%Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
 % \begin{align}
 %   A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
 % \end{align}
%and a \emph{Linear Evolutionary System} is, 
%\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
Therefore, if a system is minimal, the set of all phenotypically identical systems $\mathcal{S}(V)$ can be parameterized as,
    \begin{align} \label{eq:min_param}
      \mathcal{S}(V) := \left\{ \begin{array}{ll} V\dot{\kappa}(t) &= VA V^{-1} \kappa(t) + VB u(t) \\ \phi(t) &= C V^{-1} \kappa(t) \end{array} \right.
  \end{align}
  where $V$ is any invertible matrix and it's elements are free parameters.

Therefore varying $V$ can tweak the relationships between genes within a regulatory network, yet preserve the phenotype. Some gene networks, however, can grown or shrink, perhaps following gene duplications and deletions, and also still preserve their phenotypes. These changes cannot be described by \ref{eq:min_param}, but they are ecanpuslated by \ref{eqn:equivalence}, below. 
More generally,
we denote by $\calA_n(\mathcal{S}_0)$ the set of all $n$-dimensional systems equivalent to $\mathcal{S}_0$:
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
    \calA_n(\mathcal{S}_0) 
      &= \left\{
        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
      \right\}  \\
      &= \left\{
        (A,B,C) : C A^r B = C_0 A_0^r B_0 \; \text{for}\; 1 \le r \le n-1 
      \right\} .
  \end{aligned}
\end{equation}
Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
\plr{To make sense of this we need to say how $C$ and $B$ change with $n$.}
Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.

@@

Suppose that a set $x$ of coefficients that determine a system (this is $A$ above),
produce a phenotype $\Phi(x)$ (the time course of $\phi(t)$).
There is an optimal phenotype $\optph$,
and a set $\optx$ of ``optimal'' coefficients that produce this phenotype.
Fitness depends on distance to the optimal phenotype --
we will write the ``distance'' between phenotypes $\phi$ and $\psi$ as $\dph(\phi, \psi)$,
measured so that the fitness of an organism with coefficents $x$ is $\fitx(x) = \exp(-\dph(\Phi(x),\optph)^2)$. \js{need to add $\gamma^{2}$ here?}
We will assume that the map $\Phi$ is smooth
and that the optimal set $\optx$ is locally isomorphic to $\R^m$.
\plr{say that better}

@@

\subsection*{The rate of speciation under neutrality in allopatric populations}

@@

$F_2$s between allopatric populations will have an average fitness relative to their parents of,
  \begin{align}
    \exp\left(-\left( \frac{2\sigma_{N}}{\beta}\sqrt{\frac{T}{N_e}} \right)^{2} \right)
  \end{align}
where $\beta$ is the strength of selection against phenotypic divergence.

If the strength of selection is only a few multiples $\xi_1$ of the standing phenotypic diversity within a population $\sigma_S$, such that $\beta = \xi_1 \sigma_S$, and recombination load constrains the standing kryptotype diversity $\sigma_N$ to be only a few multiples $\xi_2$ of $\sigma_S$, then $F_2$ fitness will be,
  \begin{align}
    \exp\left(-\left( \xi \sqrt{\frac{T}{N_e}} \right)^{2} \right)
  \end{align}
  where $\xi := 2\xi_2/\xi_1$.
  This suggests that if a population contains as much phenotypic diversity as allowed by selection and if $\sigma_N$ is on the order of $\sigma_S$, reproductively isolated populations can speciate rapidly -- on timescales shorter than $N_{e}$ generations.

>>>>

\section*{Gene networks as linear dynamical systems}
%
\plr{Below we mix general language with language specific to GRNs, like switching between "internal state" and "transcription factor concentration.  
Rather than keep making comments like "(or some other system)",
we could say up top that this applies to other situations, but to make it concrete we'll talk about regulatory networks.}
%
%  \jss{Organisms' phenotypes are constructred by gene by gene by environment interactions. Here we simply define the phenotype to be the organismal temporal molecular dynamics directly under natural selection. The \emph{what}, \emph{when}, and \emph{where}, of an organism's molecules that are physiologically or otherwise relevant for survival. 
%   Thus we say that some function $\phi(t)$ is a phenotype where, 
%  \begin{align}
%    \phi(t) = \int_{0}^{\infty} h(t) u(t) dt  ,
%  \end{align}
%  and $h(t)$ is the \emph{impulse response} of the system and $u(t)$ is the \emph{input} function, both functions of time $t$. The input can be interpreted as the environment, as initial conditions, or otherwise, depending on the biological specifics under study. The phenotype is a convolution of both the impulse response and the input, as allowed by our assumption of linearity. 
%
%  Essentially the phenotype $\phi(t)$ is a consequence of an organism's specific gene by gene interactions, given by $h(t)$, reacting further with the local environment, given by $u(t)$. 
%
%  We describe the impulse response as, 
%  \begin{align}
%    h(t) = C e^{A t} B
%  \end{align}
%  where $A$ is a gene regulatory network (although this model can be generalized to other biological networks) -- a square matrix, and $B$ filters and translates the input to the system. The form of $B$ determines precisely how the state of the external environment influences the internal gene network. $C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection. 
%
%  Generally $A$ can be any real $n \times n$ matrix, $B$ any $n \times \ell$, and $C$ any $\ell \times n$ dimensional matrix. Each $i$th row of $A$ describes the \emph{cis}-regulatory module for gene $i$, and each $j$th entry, the specific regulatory influence of gene $j$ on gene $i$. 
%
%  Although $\phi(t)$ describes the phenotype given an input, $h(t)$ describes the phenotype subject only to an impulse -- an input present initially and absent immediately thereafter. Typically, a system $\Sigma$, is defined as,
%  \begin{align}
%    \Sigma = \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right.
%  \end{align}
%  Variables have the same identities as described above and $\kappa(t)$ is a vector of molecule concentrations at time $t$. Therefore the molecular concentrations at a specific time are completely determined by the input and gene by gene interactions. Lastly, a portion and/or combination of these molecules, $\phi(t)$, are ``observed'' by selection (this is in constrast to $\kappa(t)$ -- the \emph{kryptotype} -- as it is ``hidden'' from direct selection).
%}
\paragraph{Systems theory}
Organisms' phenotypes are constructred by gene by gene by environment interactions. Here we simply define the \emph{phenotype} to be the temporal molecular dynamics directly under natural selection. The \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.

@@

Finally, we can write the phenotype as a convolution of the system organization and the environment,
  \begin{align}
    \phi(t) = C e^{A t} \kappa(0) + \int_{0}^{t} C e^{A (t-s)} B u(s) ds ,
  \end{align}
where we refer to $h(t) := Ce^{A t}B$ as the system's \emph{impulse response}. 
%We will now lay out the model in more general terms.
%Suppose that the \emph{internal state} of the system
%is parameterized by the concentrations of a collection of $n$ molecular species,
%$S_1, \ldots, S_n$,
%and the vector of concentrations at time $t$ we denote $\kappa(t)=(\kappa_1(t),\ldots,\kappa_n(t))$.
%There are also $m$ ``input'' species, whose concentrations are determined
%exogenously to the system,
%and are denoted $u(t) = (u_1(t),\ldots,u_m(t))$,
%and $\ell$ ``output'' species, whose concentrations are denoted
%$\phi(t) = (\phi_1(t),\ldots,\phi_\ell(t))$.
%The output is merely a linear function of the internal state:
%\begin{align*}
%    \phi_i(t) = \sum_j C_{ij} \kappa_i(t).
%\end{align*}
%Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype},
%and sometimes in contrast refer to $\kappa$ as the \emph{kryptotype},
%as it is ``hidden'' from direct selection.
%The rate at which the $i^\text{th}$ species is produced
%is a weighted sum of the concentrations of the other species
%as well as the input:
%\begin{align*}
%    \ddt \kappa_i(t) = \sum_j A_{ij} \kappa_j(t) + \sum_r B_{ir} u_r(t) .
%\end{align*}
%In matrix notation, this is written more concisely as
%\begin{align} \label{eqn:lti_system}
%    \ddt \kappa(t) &= A \kappa(t) + B u(t) \\
%    \phi(t) &= C \kappa(t) .
%\end{align}
 \begin{example}[Oscillating gene network: cell cycle control]\label{ex:oscillator}
    Cellular division is governed by many different processes, however it is thought that its rhythm is partially controlled by oscillating gene transcription \citep{orlando2008global}. Here we consider a simplified model of oscillating gene transcription.

@@

  \paragraph{Phenotypically equivalent gene networks}

@@

Gene regulatory networks with identical phenotypes do not necessarily have identical kryptotypes.
Any linear and minimal gene network 
-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes $(n=m)$
-- has an identical impulse response and thus phenotype given identical inputs $u(t)$, up to a change of coordinates. 
\plr{"has an identical phenotype" is not what you mean to say here}
\js{edited -- does that work?}
  \begin{equation}
    \begin{aligned}
      h(t) &= C e^{A t} B 
      = C V^{-1} V e^{A t} V^{-1} V B \\
      &= C V^{-1} e^{V A V^{-1} t} V B 
      = \bar{C} e^{\bar{A} t} \bar{B}
    \end{aligned}
  \end{equation}
\plr{what are we calling these -- not "biological system" -- that's too broad}\js{I see your point, but I don't want to bog down the paper by constantly qualifying statements -- it seemsnatural to refer to biological systems, since these are, afterall dynamical systems.}
Two biological systems, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, 
have the same phenotype, for all possible inputs, if they are related by a change of coordinates. 
\plr{is this the place to say "and only if, in the minimal dimension"?}

@@

%Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
 % \begin{align}
 %   A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
 % \end{align}
%and a \emph{Linear Evolutionary System} is, 
%\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
Therefore the set of all phenotypically identical (and minimal) systems can be parameterized as, $\mathcal{S}_{m}(V) := \{VA_0V^{-1}, VB_0, C_0V \}$,
  %  \begin{align} \label{eq:min_param}
  %    \mathcal{S}(V) := \left\{ \begin{array}{ll} V\dot{\kappa}(t) &= VA V^{-1} \kappa(t) + VB u(t) \\ \phi(t) &= C V^{-1} \kappa(t) \end{array} \right.
  %\end{align}
  where $V$ is any invertible matrix and it's elements are free parameters.

Therefore varying $V$ can tweak the relationships between genes within a regulatory network, yet preserve the phenotype. Some gene networks, however, can grown or shrink, perhaps following gene duplications and deletions, and also still preserve their phenotypes. These changes cannot be described by $S(V)$, but are ecanpuslated by \ref{eqn:equivalence}, below.

More generally,
we denote by $\calA_n(\mathcal{S}_0)$ the set of all $n$-dimensional systems equivalent to $\mathcal{S}_0$:
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
    \calA_n(\mathcal{S}_0) 
      &= \left\{
        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
      \right\}  \\
      &= \left\{
        (A,B,C) : C A^r B = C_0 A_0^r B_0 \; \text{for}\; 1 \le r \le n-1 
      \right\} .
  \end{aligned}
\end{equation}
Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
\plr{To make sense of this we need to say how $C$ and $B$ change with $n$.}
Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.

@@

Suppose that a set $x$ of coefficients that determine a system (this is $A$ above),
produce a phenotype $\Phi(x)$ (the time course of $\phi(t)$).
There is an optimal phenotype $\optph$,
and a set $\optx$ of ``optimal'' coefficients that produce this phenotype ($\calA_n$ above).
Fitness depends on distance to the optimal phenotype --
we will write the ``distance'' between phenotypes $\phi$ and $\psi$ as $\dph(\phi, \psi)$,
measured so that the fitness of an organism with coefficents $x$ is $\fitx(x) = \exp(-\dph(\Phi(x),\optph)^2)$. \js{need to add $\gamma^{2}$ here?}
We will assume that the map $\Phi$ is smooth
and that the optimal set $\optx$ is locally isomorphic to $\R^m$.
\plr{say that better}

@@

\paragraph{The rate of speciation under neutrality in allopatric populations}

@@

$F_2$s between allopatric populations will suffer an average fitness drop relative to their parents of,
  \begin{align}
    \exp\left(-\left( \frac{2\sigma_{N}}{\beta}\sqrt{\frac{T}{N_e}} \right)^{2} \right) .
  \end{align}

If the strength of selection is only a few multiples $\xi_1$ of the standing phenotypic diversity within a population $\sigma_S$, such that $\beta = \xi_1 \sigma_S$, and recombination load constrains the standing phenotype-neutral diversity $\sigma_N$ to be only a few multiples $\xi_2$ of $\sigma_S$, then $F_2$ fitness drop will be,
  \begin{align}
    \exp\left(- \xi^{2} \frac{T}{N_e} \right)
  \end{align}
  where $\xi := 2\xi_2/\xi_1$.
  This suggests that if a population contains as much phenotypic diversity as allowed by selection and if $\sigma_N$ is on the order of $\sigma_S$, reproductively isolated populations can speciate rapidly -- possibly on timescales shorter than $N_{e}$ generations.
